A novel route to the spontaneous formation of porous crystals via viscoelastic phase separation

### 2015-11-19 nature/crystal_gel2.tex

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{scicite}

% \usepackage{hyperref}
\usepackage{booktabs}
\usepackage{color}

\usepackage[normalem]{ulem}

\topmargin 0.0cm
\oddsidemargin 0.2cm
\textwidth 16cm 
\textheight 21cm
\footskip 1.0cm

\newenvironment{sciabstract}{%
\begin{quote} \bf}
{\end{quote}}

\newcounter{lastnote}
\newenvironment{scilastnote}{%
\setcounter{lastnote}{\value{enumiv}}%
\addtocounter{lastnote}{+1}%
\begin{list}%
{\arabic{lastnote}.}
{\setlength{\leftmargin}{.22in}}
{\setlength{\labelsep}{.5em}}}
{\end{list}}

\title{A novel route to crystallization via viscoelastic phase separation}

\author{Hideyo Tsurusawa$^1$, John Russo$^1$, Mathieu Leocmach$^{2}$ and Hajime Tanaka$^{1,\ast}$ 
\\
\normalsize{$^1$Institute of Industrial Science, University of Tokyo,}\\
\normalsize{4-6-1 Komaba, Meguro-ku, Tokyo 153-8505, Japan}\\
\normalsize{$^2$Institut Lumière Matière, CNRS UMR 5306, Université Claude Bernard Lyon 1, }\\
\normalsize{Université de Lyon, Lyon, 69622 Villeurbanne Cedex, France}\\
\normalsize{$^\ast$Corresponding author. E-mail: tanaka@iis.u-tokyo.ac.jp}}

%\date{Received}

\begin{document}

\baselineskip24pt

% Make the title.

\maketitle
\vspace{-1.5cm}
\begin{sciabstract}
With confocal microscopy measurements we find a novel crystallization pathway of a colloid-polymer mixture undergoing viscoelastic gas-liquid phase separation. We show that crystallization leads to a network structure different from the original phase separation pattern. Crystals nucleate inside the liquid network, but grow past it by direct condensation of the gas phase on their surface, driving evaporation of nearby liquid. This process represents the colloidal analogue of the Bergeron process, i.e., ice crystals formation in mixed phase clouds and is at the origin of rain. Our finding gives us full experimental access to the kinetic pathway of this phenomenon important for climate science. We argue that similar states can form as the result of the gas-liquid phase separation of crystallizable components, such as monoatomic and molecular systems, if dynamics of the liquid phase is slow enough to induce viscoelastic phase separation, but fast enough to prevent immediate vitrification.
\end{sciabstract}

@@

Colloidal suspensions offer a system where the crystallization process in a mixed-phase environment can be observed with single-particle
resolution, and at the same time scales over which nucleation takes place. In colloids with short-range attractions the liquid-gas transition 
becomes metastable with respect to crystallization~\cite{anderson2002insights,lekkerkerker2011colloids}, and can form gels
by viscoelastic phase separation \cite{tanaka1999colloid,tanaka2000viscoelastic}, with the coexistence of a dense liquid phase with slow dynamics and a dilute gas phase with fast dynamics. 
This difference in the viscoelastic properties between the two phases allows the system to form a space-spanning 
network structure of the liquid phase, even if it is the minority phase. The network is fractal and its dynamic structure factor shows a stretched-exponential decay to a finite
plateau~\cite{krall1998internal,solomon2001dynamic,romer2000sol}. 
During phase separation, the density of the liquid phase increases towards the glass-transition density, leading to slow glassy dynamics.
At the glass transition point, the percolated network structure is dynamically stabilised by vitrification 
of the dense liquid phase~\cite{pusey1993dynamics,ilett1995phase,verhaegh1997transient,tanaka1999colloid,foffi2002,buzzaccaro2007sticky,zaccarelli2007,lu2008gelation,zaccarelli2008gelation,testard2011}.  
More precisely, the dynamical stabilization is due to percolation of locally favored structures, which are locally stable non-crystalline structures \cite{royall2008g}.

@@

Unlike the above standard scenario of colloidal gelation, it is known that phase separation can also be arrested by crystallization, if the process takes place below the melting point of a component of a mixture~\cite{tanaka1985new}. Indeed the possibility of a different class of gels, which are stabilized by crystallization, 
was suggested by numerical simulations~\cite{fortini2008crystallization,perez2011pathways} and observed in microgravity experiments \cite{sabin2012}. 
For colloidal systems, we can in principle access structural evolution in real time at a single-particle level; however, there has so far been no confocal microscopy  
study on the dynamical process of crystal-gel formation. 
For confocal microscopy experiments on colloid gelation, colloid-polymer mixtures have so far been used as a model system. 
The main difficulty consists in the preparation protocol of such a colloid-polymer mixture, where before observation of its dynamics shaking or applying shear is necessary to prepare an initial homogeneous state. This introduces turbulent flows at the beginning of the process, and also does not allow the
observation of the initial stages of crystallization. This initial perturbation may even alter the selection of the final non-equilibrium arrested state.  
In our experiments we instead succeed in building a novel protocol that allows the time evolution of the system to be observed directly
under the confocal microscope from the very early stages and in a quiescent situation without introducing fluid flows.

In this Report, we show that, for low polymer concentration, the system can undergo a novel crystallization pathway that leads to the formation
of crystalline droplets. These droplets originate from crystal seeds that form inside the liquid branches of the spinodal aggregate of network morphology, but then grow by direct condensation from the gas phase. The process is analogous to the Bergeron process of ice formation in mixed phase clouds, where ice droplets grow at the
expense of the supercooled liquid droplets due to their lower saturated vapor pressure. 
This is speculated to be the primary mechanism for the formation of raindrops. 
The mechanism found here is in contrast to the standard crystal-gel scenario, where crystals nucleate and grow inside the dense branches of the gel, becoming dynamically arrested. We also observe that increasing the strength of the interparticle attraction
(by increasing the polymer concentration), instead of forming crystal gels, the system undergoes the classical gelation process, in which
phase separation is interrupted by vitrification~\cite{verhaegh1997transient,tanaka1999colloid,Royall2007,lu2008gelation}.

In our experimental setup, the sample cell is put in contact with a reservoir of a salt solution through a semi-permeable membrane (see Methods 
on the details of our experiments).
At $t=0$, the increase in the concentration of salt ions within a few Brownian times ($\tau_B$) screens the Coulombic repulsion between colloidal particles,
which are then subject to attractive depletion forces, making the system thermodynamically unstable and leading to phase separation. 
%In the following we discuss all the stages of the formation of the crystal gels: phase separation, stress-driven agein and crystallization.
The experimental data are taken at two different volume fractions ($\phi\approx 0.13$ and $\phi\approx 0.30$) and for different values
of the polymer concentration, $c_p=0.38,0.48,0.57,1.07$ mg/g for $\phi\approx 0.30$ and $c_p=0.82,1.36$ mg/g for $\phi\approx 0.13$.
In Fig.~\ref{fig:phase_separation}A, we superimpose the state points with the phase diagram, which we calculated from generalised free volume theory~\cite{Fleer2008} using the experimental values for the size of the colloids, the radius of gyration of the polymer, and the polymer concentration (see Supplementary Information).
For each state point, scans were made at early times (every $10$~s, i.e. $4\tau_B$) and at the late stage (every $30$~s) of the gelation process.

%
%\subsection*{Phase separation}
%
%In Fig.~\ref{fig:gel}a we show the cluster size distribution for all state points at their gelation point. The decay of the distribution
%follows different power laws $n(s)\sim s^\tau$, where $\tau$ is the Fisher exponent \cite{family2012kinetics}.
%For state points with $\phi\approx 0.30$ the decay is well described by the three dimensional (3D) random percolation exponent $\tau=2.18$ \cite{family2012kinetics}.
%For state points with $\phi\approx 0.13$, on the other hand, the decay appears to be slower, close to the diffusion limited cluster aggregation (DLCA) 
%exponent $\tau=1.8$ \cite{family2012kinetics}. While it is plausible that gelation at lower volume fraction occurs through cluster aggregation, especially given 
%the high polymer concentration of these state points, we have to warn that the cluster size distribution is very sensitive to
%experimental conditions and statistical noise (which is high in our case since the distributions are extracted from a single frame).
%Other measures of fractal dimensions (see below) will in fact suggest that also the $\phi\approx 0.13$ states belong to the random gel universality class.
%
%Next we show in Fig.~\ref{fig:gel}b the radius of gyration $R_{\rm g}$ for the clusters of size $s$ at the gel point. All clusters, irrespective
% of the state point, follow the power-law: $R_{\rm g}\sim s^{2/D}$, with $D=2.53$ the fractal dimension in the random gelation
% universality class \cite{family2012kinetics}. So far, we have thus proven that, in the initial stages of phase separation, the liquid branch is
% formed by colloidal particles that aggregate to form a random gel.
% 
% Even after a gel is formed, the system keeps evolving.
% However, since a gel has an elasticity due to its space-spanning connectivity, the process should be markedly different from 
% ordinary fluid phase separation. Now we consider this unique feature of the coarsening of a gel network.  
% 
% \subsection*{Stress-driven ageing}
% 
% Previous simulation results~\cite{tanaka2000,tanaka2007spontaneous,furukawa2010key} have suggested that hydrodynamic interactions play
% a key role in the process of colloidal gelation. Without hydrodynamic interactions, particles have the tendency to
% aggregate in compact structures and subsequently form thick network structures. This also limits the possibility to
% form arrested gel states only at relatively high volume fractions. With hydrodynamic interactions, particles first form
% a transient gel even at very low volume fractions, and only later they increase the number of
% nearest neighbours to minimize the energy of the structure. 
% Thus, hydrodynamic interactions lead to the formation of gels that are very far from equilibrium and under a strong thermodynamic driving force 
% towards more stable compact structures. 
% The resulting transition from open to more compact networks occurs only through
% the breaking of the bonds that have accumulated more stress. The process was called stress-driven ageing~\cite{tanaka2007spontaneous}, 
% which is characteristic of viscoelastic phase separation \cite{tanaka2000viscoelastic}. 
% Below we provide the first experimental proof of this process.
% 
% In Fig.~\ref{fig:stress}a, we plot the fraction of liquid and gas particles as a function of time during phase separation
% at the state points of $\phi\approx 0.13$. We observe that at high polymer concentration ($c_p=1.36$, dashed lines) both the
% liquid and gas branch change monotonically with time as the phase separation proceeds. At low polymer concentration
% ($c_p=0.82$, continuous lines), on the other hand, we find a non-monotonic evolution of both the liquid and gas branches, which is signalling
% a structural reorganization of the network. To prove that this is indeed a stress-driven ageing process, in Fig.~\ref{fig:stress}b 
% we plot the average number of bond-breaking events for the low-polymer concentration state point. The figure clearly shows that
% the network restructuring follows a peak in the number of bond-breaking events. We have thus found direct evidence of the
% stress-driven ageing process in colloidal gels that is a direct consequence of the formation of a thin network with the help of hydrodynamic interactions in the first stages
% of the aggregation process.
% 
% In Fig.~\ref{fig:stress}c we plot the time evolution of the structure factor, $S(q)$, for the state point
% undergoing network reorganization $\phi\approx 0.13$ and $c_p=0.82$. The time of network reorganization as found in Fig.~\ref{fig:stress}a and b
% corresponds to a steep increase of the low wavenumber scattering from the network, that signals the coarsening of the network structure. 
% 
% The same conclusions also hold for the state point at $\phi\approx 0.30$. We observe network restructuring only for the state point
% at lower polymer concentration, $c_p=0.38$. Interestingly, this is also the only state point that undergoes crystallization,
% meaning that the coarsening of the network of the dense liquid phase and the resulting formation of a wider liquid branch is a necessary condition for crystallization.

As described above and also in Methods, salt injection initiates liquid-gas phase separation of the colloidal suspension. 
All samples share the same early stages of spinodal decomposition. 
Due to strong dynamical asymmetry between colloids and the solvent \cite{tanaka1999colloid}, the colloidal particles start aggregating, 
eventually forming a percolating network. Thus, a dense network (liquid) coexists with freely diffusing monomers (gas).
In Fig.~\ref{fig:phase_separation}A, the state points which did not phase separate are indicated with open triangles, while
the other symbols indicate samples undergoing phase separation. The (black) continuous line is the gas spinodal computed from free volume theory,
and it is in good agreement with our experimental points (the small discrepancy at small colloidal volume fractions 
is common in free volume theory)~\cite{Royall2007,lu2008gelation}. 
The topology of the network is similar to all our samples at the initial stages. 
This is shown for example in Fig.~\ref{fig:phase_separation}B, where we plot the radius of gyration of colloidal clusters ($R_g$)
as a function of the cluster size ($s$) at the percolation point. Two particles belong to the same cluster if their distance is
within the first peak of the radial distribution function. All samples show the same scaling law, $R^2_{\rm g}\sim s^{2/D}$, which is compatible with
the random gelation universality class exponent ($D=2.53$) as shown by the dashed line in Fig.~\ref{fig:phase_separation}B
(we do not imply that the formation of the network follows the random gelation universality class, just that our
fractal dimension is compatible with it).

While sharing similar structural properties at early times, the samples show significantly different behaviors in the later stages of the phase separation process.
To examine this in more detail, we plot in Fig.~\ref{fig:phase_separation}C the structure factor for the final stages of the gelation process for all state points.
Calculation of the structure factor is done with the Hanning window function, to minimize boundary effects.
The figure shows that at low wavenumber $q$, the structure factor displays fractal scaling compatible with Guinier law, $S(q)\sim q^{-D}$.
But a difference in the fractal dimension $D$ between the $\phi\approx 0.30$ and $c_p=0.38$ mg/g state point and other state points starts to become visible. In fact, while states
with high polymer concentration retain the exponent $D=2.5(3)$, 
which is the random gel universality class exponent, as we also observed in the early stages of the gelation process (see Fig.~\ref{fig:phase_separation}B),
the state point with low polymer concentration ($\phi\approx 0.30$ and $c_p=0.38$ mg/g) displays the biggest deviation from that exponent. In the figure we also
plot the exponent $D=3$, which corresponds to volume growth, which we will show in the following analysis to be the correct exponent for this state. 
The difference is due to the onset of crystallization in the low-polymer concentration sample. This is already evident in the high $q$ behavior of
the structure factors (Fig.~\ref{fig:phase_separation}C), where the diffraction planes appear as sharper peaks for $\phi\approx 0.30$ and $c_p=0.38$ mg/g.

To gain additional insight, in Fig.~\ref{fig:phase_separation}D we plot the average local volume fraction, computed
through Voronoi diagrams, as a function of the number of neighbors. For all state points, we can see that the average volume fraction of a colloidal
particle increases with its number of neighbors. Approaching close packing (12 neighbors), we clearly see two families of curves. The state point
at $\phi\approx 0.30$ and $c_p=0.38$ mg/g reaches close packing at a volume fraction of around $\sim 68\%$, which is in reasonable agreement with the
volume fraction of the stable crystal phase~\cite{ilett1995phase}. All other state points, instead, reach close packing at a volume fraction $\sim 64\%$, which is
indeed close to the volume fraction of the attractive glass state~\cite{pham2002multiple}. 
%(\textbf{the absolute values of these volume fraction are not totally convincing, probably because of a not optimal estimate of the size of the colloids. We should run the code of Mathieau with size detection}). 
This result clearly indicates that the state point at $\phi\approx 0.30$ and $c_p=0.38$ mg/g follows a different arrest mechanism, in which phase separation
is arrested by crystallization and not by glassiness. Although our estimation of the absolute volume fraction may involves systematic errors of a few \%, the trend should be robust~\cite{Poon2012}.

% 
%  REINTREGRARE
% while the crystallizing state point shows a scaling compatible with $D=3$, which
% signals compact crystalline nuclei. At high $q$, the first signals from diffraction planes are also evident in the $\phi\approx 0.30$ and $c_p=0.38$ state
% point. This suggests that the space filling network is composed of crystalline aggregates, differently from the fractal branches which characterize the
% spinodal liquid. Also this suggests that the growth mechanism of the crystal is not simply due to filling of the spinodal liquid network, as was
% instead observed in Ref.~\cite{fortini2008crystallization,perez2011pathways} for brownian dynamics simulations.

@@

To confirm the formation of crystalline seeds, we adopt bond orientational analysis to detect which particles are in crystalline environments~\cite{russo2013interplay}. 
With this analysis we reveal that crystallization events indeed start occurring for the state point $\phi\approx 0.30$ and $c_p=0.38$ mg/g, which has the lowest
polymer concentration of all state points. This is shown in Fig.~\ref{fig:crystals}A, where the black line indicates the average size of the crystallites, 
while the red line indicates the average number of crystallites as a function of time. The number of crystallites first rapidly increases as nucleation events
start occurring inside the liquid network, but eventually starts decreasing as the different crystallites grow and merge with each other.
%The average size of the crystals shows an interesting linear growth regime for $100<t<600$. This linear growth regime goes past the time
%of coalescence of nuclei, so a first speculation is that it could be due to the compensation of the following two effects:
%\begin{itemize}
% \item diffusive growth of the nuclei, which scales like $t^{1/2}$
% \item coalescence growth, which should scale faster than $t$, maybe $t^2?$
%\end{itemize}
All other state points show only negligible signs of crystallization. In particular, for state points with $\phi\approx 0.30$, increasing
polymer concentration drastically reduces the amount of crystals. This is in agreement with the
idea of enhanced crystallization rates near metastable critical points~\cite{ten1997enhancement,olmsted1998spinodal} and the
results of Refs.~\cite{fortini2008crystallization,perez2011pathways},
which speculated two different arrest mechanisms: crystallization at low polymer concentration, and dynamic arrest at high polymer concentrations.

>>>>

\documentclass[preprint,amsmath,amssymb,superscriptaddress]{revtex4-1}
\usepackage{graphicx,amsmath}
\usepackage{color}
\begin{document}
\draft

\title{A novel route to the spontaneous formation of porous crystals via viscoelastic phase separation}

\author{Hideyo Tsurusawa} 
\affiliation{Institute of Industrial Science, University of Tokyo, 4-6-1 Komaba, Meguro-ku, Tokyo 153-8505, Japan}
\author{John Russo}
\affiliation{Institute of Industrial Science, University of Tokyo, 4-6-1 Komaba, Meguro-ku, Tokyo 153-8505, Japan}
\author{Mathieu Leocmach}
\affiliation{Institut Lumière Matière, CNRS UMR 5306, Université Claude Bernard Lyon 1, Université de Lyon, Lyon, 69622 Villeurbanne Cedex, France}
\author{Hajime Tanaka}
\affiliation{Institute of Industrial Science, University of Tokyo, 4-6-1 Komaba, Meguro-ku, Tokyo 153-8505, Japan}

\date{Received \today}

\begin{abstract}
{\bf
Viscoelastic phase separation of colloidal suspensions can be interrupted either by the glass transition, with the formation of colloidal gels~\cite{poon2002,zaccarelli2007,piazza1994phase,verhaegh1997transient,tanaka1999colloid,lu2008gelation}, or by crystallization, which results in the formation of a crystalline network~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways,sabin2012}. But while the former process has been studied in detail, the latter has never been observed experimentally at a single-particle level and it is thus lacking a microscopic understanding. With a new confocal microscopy protocol, we unveil a novel crystallization pathway to sponge-like porous structures in a colloid-polymer mixture undergoing viscoelastic gas-liquid phase separation. We show that crystallization leads to a network structure different from the original phase separation pattern. Crystals nucleate inside the liquid network, but grow past it by direct condensation of the gas phase on their surface, driving evaporation of nearby liquid. This process represents the colloidal analogue of the Bergeron process~\cite{glickman2000glossary}, i.e., ice crystals formation in mixed phase clouds and is at the origin of rain. Our finding gives us full experimental access to the kinetic pathway of this poorly understood phenomenon important for climate science. \textcolor{blue}{The key is complex dynamic interplay involving the three phases upon simultaneous occurrence of two non-equilibrium phenomena, phase separation and crystallization. We argue that similar crystal-gel states can be formed as the result of the gas-liquid phase separation of any crystallizable components, such as monoatomic and molecular systems, if dynamics of the liquid phase is slow enough to induce viscoelastic phase separation, but fast enough to prevent immediate vitrification. This novel mechanism not only realizes single-step spontaneous formation of nano-porous crystals of metals and semiconductors without dealloying \cite{erlebacher2001evolution}, which may be important for catalytic, optical, sensing, and filtration applications \cite{ding2004nanoporous,ding2009nanoporous, wittstock2010nanoporous}, but also sheds new light on crystal network formation in more complex systems such as magma \cite{philpotts1998role}, biominerals \cite{rousseau2005multiscale}, and foods \cite{deman1987fat}.} }
\end{abstract}

@@

Colloidal suspensions offer a system where the crystallization process in a mixed-phase environment can be observed with single-particle
resolution, and at the same time scales over which nucleation takes place. In colloids with short-range attractions the liquid-gas transition 
becomes metastable with respect to crystallization~\cite{anderson2002insights,lekkerkerker2011colloids}, and can form gels
by viscoelastic phase separation \cite{tanaka1999colloid,tanaka2000viscoelastic}, with the coexistence of a dense liquid phase with slow dynamics and a dilute gas phase with fast dynamics. 
This difference in the viscoelastic properties between the two phases allows the system to form a space-spanning 
network structure of the liquid phase, even if it is the minority phase. The network is fractal and its dynamic structure factor shows a stretched-exponential decay to a finite
plateau~\cite{krall1998internal,solomon2001dynamic,romer2000sol}. 
During phase separation, the density of the liquid phase increases towards the glass-transition density, leading to slow glassy dynamics.
At the glass transition point, the percolated network structure is dynamically stabilised by vitrification 
of the dense liquid phase~\cite{pusey1993dynamics,piazza1994phase,ilett1995phase,verhaegh1997transient,tanaka1999colloid,foffi2002,buzzaccaro2007sticky,zaccarelli2007,lu2008gelation,zaccarelli2008gelation,testard2011}.  
More precisely, the dynamical stabilization is due to percolation of locally favoured structures, which are locally stable non-crystalline structures \cite{royall2008g}.

@@

Unlike the above standard scenario of colloidal gelation, it is known that phase separation can also be arrested by crystallization, if the process takes place below the melting point of a component of a mixture~\cite{tanaka1985new}. Indeed the possibility of a different class of gels, which are stabilized by crystallization, 
was suggested by numerical simulations~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways} and observed in microgravity experiments \cite{sabin2012}. 
For colloidal systems, we can in principle access structural evolution in real time at a single-particle level; however, there has so far been no confocal microscopy  
study on the dynamical process of crystal-gel formation. 
For confocal microscopy experiments on colloid gelation, colloid-polymer mixtures have so far been used as a model system. 
The main difficulty consists in the preparation protocol of such a colloid-polymer mixture, where before observation of its dynamics shaking or applying shear is necessary to prepare an initial homogeneous state. This introduces turbulent flows at the beginning of the process, and also does not allow the
observation of the initial stages of crystallization. This initial perturbation may even alter the selection of the final non-equilibrium arrested state.  
In our experiments we instead succeed in building a novel protocol that allows the time evolution of the system to be observed directly
under the confocal microscope from the very early stages and in a quiescent situation without introducing fluid flows.

In this Letter, we show that, for low polymer concentration, the system can undergo a novel crystallization pathway that leads to the formation
of crystalline droplets. These droplets originate from crystal seeds that form inside the liquid branches of the spinodal aggregate of network morphology, but then grow by direct condensation from the gas phase. The process is analogous to the Bergeron process of ice formation in mixed phase clouds, where ice droplets grow at the
expense of the supercooled liquid droplets due to their lower saturated vapour pressure. 
This is speculated to be the primary mechanism for the formation of raindrops. 
The mechanism found here is in contrast to the standard crystal-gel scenario, where crystals nucleate and grow inside the dense branches of the gel, becoming dynamically arrested. We also observe that increasing the strength of the interparticle attraction
(by increasing the polymer concentration), instead of forming crystal gels, the system undergoes the classical gelation process, in which
phase separation is interrupted by vitrification~\cite{verhaegh1997transient,tanaka1999colloid,Royall2007,lu2008gelation}.

In our experimental setup, the sample cell is put in contact with a reservoir of a salt solution through a semi-permeable membrane (see Methods 
on the details of our experiments).
At $t=0$, the increase in the concentration of salt ions within a few Brownian times ($\tau_B$) screens the Coulombic repulsion between colloidal particles,
which are then subject to attractive depletion forces, making the system thermodynamically unstable and leading to phase separation. 
%In the following we discuss all the stages of the formation of the crystal gels: phase separation, stress-driven agein and crystallization.
The experimental data are taken at two different volume fractions ($\phi\approx 0.13$ and $\phi\approx 0.30$) and for different values
of the polymer concentration, $c_p=0.38,0.48,0.57,1.07$ mg/g for $\phi\approx 0.30$ and $c_p=0.82,1.36$ mg/g for $\phi\approx 0.13$.
In Fig.~\ref{fig:phase_separation}a, we superimpose the state points with the phase diagram, which we calculated from generalised free volume theory~\cite{Fleer2008} using the experimental values for the size of the colloids, the radius of gyration of the polymer, and the polymer concentration (see Methods).
For each state point, scans were made at early times (every $10$~s, i.e. $4\tau_B$) and at the late stage (every $30$~s) of the gelation process.

%
%\subsection*{Phase separation}
%
%In Fig.~\ref{fig:gel}a we show the cluster size distribution for all state points at their gelation point. The decay of the distribution
%follows different power laws $n(s)\sim s^\tau$, where $\tau$ is the Fisher exponent \cite{family2012kinetics}.
%For state points with $\phi\approx 0.30$ the decay is well described by the three dimensional (3D) random percolation exponent $\tau=2.18$ \cite{family2012kinetics}.
%For state points with $\phi\approx 0.13$, on the other hand, the decay appears to be slower, close to the diffusion limited cluster aggregation (DLCA) 
%exponent $\tau=1.8$ \cite{family2012kinetics}. While it is plausible that gelation at lower volume fraction occurs through cluster aggregation, especially given 
%the high polymer concentration of these state points, we have to warn that the cluster size distribution is very sensitive to
%experimental conditions and statistical noise (which is high in our case since the distributions are extracted from a single frame).
%Other measures of fractal dimensions (see below) will in fact suggest that also the $\phi\approx 0.13$ states belong to the random gel universality class.
%
%Next we show in Fig.~\ref{fig:gel}b the radius of gyration $R_{\rm g}$ for the clusters of size $s$ at the gel point. All clusters, irrespective
% of the state point, follow the power-law: $R_{\rm g}\sim s^{2/D}$, with $D=2.53$ the fractal dimension in the random gelation
% universality class \cite{family2012kinetics}. So far, we have thus proven that, in the initial stages of phase separation, the liquid branch is
% formed by colloidal particles that aggregate to form a random gel.
% 
% Even after a gel is formed, the system keeps evolving.
% However, since a gel has an elasticity due to its space-spanning connectivity, the process should be markedly different from 
% ordinary fluid phase separation. Now we consider this unique feature of the coarsening of a gel network.  
% 
% \subsection*{Stress-driven ageing}
%

As described above and also in Methods, salt injection initiates liquid-gas phase separation of the colloidal suspension. 
All samples share the same early stages of spinodal decomposition. 
A typical phase ordering process observed at $\phi\approx 0.30$ and $c_p=0.38$ mg/g can be seen in Supplementary Movies 1 and 2. 
Due to strong dynamical asymmetry between colloids and the solvent \cite{tanaka1999colloid}, the colloidal particles start aggregating, 
eventually forming a percolating network. Thus, a dense network (liquid) coexists with freely diffusing monomers (gas).
In Fig.~\ref{fig:phase_separation}a, the state points which did not phase separate are indicated with open triangles, while
the other symbols indicate samples undergoing phase separation. The (black) continuous line is the gas spinodal computed from free volume theory,
and it is in good agreement with our experimental points (the small discrepancy at small colloidal volume fractions 
is common in free volume theory)~\cite{Royall2007,lu2008gelation}. 
The topology of the network is similar to all our samples at the initial stages. 
This is shown for example in Fig.~\ref{fig:phase_separation}b, where we plot the radius of gyration of colloidal clusters ($R_g$)
as a function of the cluster size ($s$) at the percolation point. Two particles belong to the same cluster if their distance is
within the first peak of the radial distribution function. All samples show the same scaling law, $R^2_{\rm g}\sim s^{2/D}$, which is compatible with
the random gelation universality class exponent ($D=2.53$) as shown by the dashed line in Fig.~\ref{fig:phase_separation}b
(we do not imply that the formation of the network follows the random gelation universality class, just that our
fractal dimension is compatible with it).

%%%%
%Previous simulation results~\cite{tanaka2000,tanaka2007spontaneous,furukawa2010key} have suggested that hydrodynamic interactions play a key role in the process of colloidal gelation. Without hydrodynamic interactions, particles have the tendency to aggregate in compact structures and subsequently form thick network structures. This also limits the possibility to form arrested gel states only at relatively high volume fractions. With hydrodynamic interactions, particles first form a transient gel even at very low volume fractions, and only later they increase the number of nearest neighbours to minimize the energy of the structure. Thus, hydrodynamic interactions lead to the formation of gels that are very far from equilibrium and under a strong thermodynamic driving force towards more stable compact structures. The resulting transition from open to more compact networks occurs only through the breaking of the bonds that have accumulated more stress. The process was called stress-driven ageing~\cite{tanaka2007spontaneous}, which is characteristic of viscoelastic phase separation \cite{tanaka2000viscoelastic}. Below we provide the first experimental proof of this process.

%In Fig.~\ref{fig:stress}a, we plot the fraction of liquid and gas particles as a function of time during phase separation at the state points of $\phi\approx 0.13$. We observe that at high polymer concentration ($c_p=1.36$, dashed lines) both the liquid and gas branch change monotonically with time as the phase separation proceeds. At low polymer concentration ($c_p=0.82$, continuous lines), on the other hand, we find a non-monotonic evolution of both the liquid and gas branches, which is signalling a structural reorganization of the network. To prove that this is indeed a stress-driven ageing process, in Fig.~\ref{fig:stress}b we plot the average number of bond-breaking events for the low-polymer concentration state point. The figure clearly shows that the network restructuring follows a peak in the number of bond-breaking events. We have thus found direct evidence of the stress-driven ageing process in colloidal gels that is a direct consequence of the formation of a thin network with the help of hydrodynamic interactions in the first stages of the aggregation process.

%In Fig.~\ref{fig:stress}c we plot the time evolution of the structure factor, $S(q)$, for the state point undergoing network reorganization $\phi\approx 0.13$ and $c_p=0.82$. The time of network reorganization as found in Fig.~\ref{fig:stress}a and b corresponds to a steep increase of the low wavenumber scattering from the network, that signals the coarsening of the network structure.

%The same conclusions also hold for the state point at $\phi\approx 0.30$. We observe network restructuring only for the state point at lower polymer concentration, $c_p=0.38$. Interestingly, this is also the only state point that undergoes crystallization, meaning that the coarsening of the network of the dense liquid phase and the resulting formation of a wider liquid branch is a necessary condition for crystallization.

%%%

While sharing similar structural properties at early times, the samples show significantly different behaviours in the later stages of the phase separation process.
To examine this in more detail, we plot in Fig.~\ref{fig:phase_separation}c the structure factor for the final stages of the gelation process for all state points.
Calculation of the structure factor is done with the Hanning window function, to minimize boundary effects.
The figure shows that at low wavenumber $q$, the structure factor displays fractal scaling compatible with Guinier law, $S(q)\sim q^{-D}$.
But a difference in the fractal dimension $D$ between the $\phi\approx 0.30$ and $c_p=0.38$ mg/g state point and other state points starts to become visible. In fact, while states
with high polymer concentration retain the exponent $D=2.5(3)$, 
which is the random gel universality class exponent, as we also observed in the early stages of the gelation process (see Fig.~\ref{fig:phase_separation}b),
the state point with low polymer concentration ($\phi\approx 0.30$ and $c_p=0.38$ mg/g) displays the biggest deviation from that exponent. In the figure we also
plot the exponent $D=3$, which corresponds to volume growth, which we will show in the following analysis to be the correct exponent for this state. 
The difference is due to the onset of crystallization in the low-polymer concentration sample. This is already evident in the high $q$ behaviour of
the structure factors (Fig.~\ref{fig:phase_separation}c), where the diffraction planes appear as sharper peaks for $\phi\approx 0.30$ and $c_p=0.38$ mg/g.

To gain additional insight, in Fig.~\ref{fig:phase_separation}d we plot the average local volume fraction, computed
through Voronoi diagrams, as a function of the number of neighbours. For all state points, we can see that the average volume fraction of a colloidal
particle increases with its number of neighbours. Approaching close packing (12 neighbours), we clearly see two families of curves. The state point
at $\phi\approx 0.30$ and $c_p=0.38$ mg/g reaches close packing at a volume fraction of around $\sim 68\%$, which is in reasonable agreement with the
volume fraction of the stable crystal phase~\cite{ilett1995phase}. All other state points, instead, reach close packing at a volume fraction $\sim 64\%$, which is
indeed close to the volume fraction of the attractive glass state~\cite{pham2002multiple}. 
%(\textbf{the absolute values of these volume fraction are not totally convincing, probably because of a not optimal estimate of the size of the colloids. We should run the code of Mathieau with size detection}). 
This result clearly indicates that the state point at $\phi\approx 0.30$ and $c_p=0.38$ mg/g follows a different arrest mechanism, in which phase separation
is arrested by crystallization and not by glassiness. Although our estimation of the absolute volume fraction may involves systematic errors of a few \%, the trend should be robust~\cite{Poon2012}.

% 
%  REINTREGRARE
% while the crystallizing state point shows a scaling compatible with $D=3$, which
% signals compact crystalline nuclei. At high $q$, the first signals from diffraction planes are also evident in the $\phi\approx 0.30$ and $c_p=0.38$ state
% point. This suggests that the space filling network is composed of crystalline aggregates, differently from the fractal branches which characterize the
% spinodal liquid. Also this suggests that the growth mechanism of the crystal is not simply due to filling of the spinodal liquid network, as was
% instead observed in Ref.~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways} for brownian dynamics simulations.

@@

To confirm the formation of crystalline seeds, we adopt bond orientational analysis to detect which particles are in crystalline environments~\cite{russo2013interplay}. 
With this analysis we reveal that crystallization events indeed start occurring for the state point $\phi\approx 0.30$ and $c_p=0.38$ mg/g, which has the lowest
polymer concentration of all state points. This is shown in Fig.~\ref{fig:crystals}a, where the black line indicates the average size of the crystallites, 
while the red line indicates the average number of crystallites as a function of time. The number of crystallites first rapidly increases as nucleation events
start occurring inside the liquid network, but eventually starts decreasing as the different crystallites grow and merge with each other.
%The average size of the crystals shows an interesting linear growth regime for $100<t<600$. This linear growth regime goes past the time
%of coalescence of nuclei, so a first speculation is that it could be due to the compensation of the following two effects:
%\begin{itemize}
% \item diffusive growth of the nuclei, which scales like $t^{1/2}$
% \item coalescence growth, which should scale faster than $t$, maybe $t^2?$
%\end{itemize}
All other state points show only negligible signs of crystallization. In particular, for state points with $\phi\approx 0.30$, increasing
polymer concentration drastically reduces the amount of crystals. This is in agreement with the
idea of enhanced crystallization rates near metastable critical points~\cite{ten1997enhancement,olmsted1998spinodal} and the
results of Refs.~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways},
which speculated two different arrest mechanisms: crystallization at low polymer concentration, and dynamic arrest at high polymer concentrations.
